---
title: "2_descriptives"
author: "Milena Costa"
date: "2025-11-04"
output: pdf_document
---

```{r setup, include=FALSE}

source(here::here("scripts","_common.R"), local = knitr::knit_global())

df_wide <- readRDS(p_processed("df_wide.rds"))
df_long <- readRDS(p_processed("df_long.rds"))


```


# Helper functions

```{r}

# Color palettes consistent across plots

pal_target <- c(
"self"      = "#E15759",
"climate"   = "#59A14F",
"prosocial" = "#4E79A7"
)



group_labels <- c(
  "control"       = "Control Group",
  "positive_norm" = "Positive Norm",
  "negative_norm" = "Negative Norm"
)


```

# Descriptive Statistics

## 1. Sample Characteristics

Currently we do not have demographic information, thus the r chunk below does not run. Should any of this information be available in the future, the following entries should be removed from the r chunk below.

```{=latex}
\begin{verbatim}
 include = FALSE, eval = FALSE
\end{verbatim}
```


```{r, include = FALSE, eval = FALSE}

# Basic demographics

# Overall sample
nrow(df_wide)  # Total N
table(df_wide$gender)
mean(df_wide$age, na.rm = TRUE)
sd(df_wide$age, na.rm = TRUE)
range(df_wide$age, na.rm = TRUE)

# By school
df_wide |> 
  group_by(school) |> 
  summarise(
    n = n(),
    mean_age = mean(age, na.rm = TRUE),
    sd_age = sd(age, na.rm = TRUE)
  )

# By experimental group
df_wide |> 
  group_by(group) |> 
  summarise(
    n = n(),
    mean_age = mean(age, na.rm = TRUE),
    sd_age = sd(age, na.rm = TRUE),
    n_female = sum(gender == "female", na.rm = TRUE),
    pct_female = mean(gender == "female", na.rm = TRUE) * 100
  )

```

## 2. Primary Outcome: Choice Behaviour

```{r}
# Overall choice rates

desc_overall <- df_long |>
summarise(
n_trials = n(),
n_valid = sum(!is.na(choice)),
n_missing = sum(is.na(choice)),
mean_choice = mean(choice, na.rm = TRUE),
sd_choice = sd(choice, na.rm = TRUE)
)

# By target type
desc_by_target <- df_long |>
group_by(target) |>
summarise(
n_trials = n(),
n_valid = sum(!is.na(choice)),
n_missing = sum(is.na(choice)),
mean_choice = mean(choice, na.rm = TRUE),
sd_choice = sd(choice, na.rm = TRUE),
.groups = "drop"
)

# By group and block 

desc_by_group_block <- df_long |>
group_by(group, block) |>
summarise(
n_trials = n(),
n_valid = sum(!is.na(choice)),
n_missing = sum(is.na(choice)),
mean_choice = mean(choice, na.rm = TRUE),
sd_choice = sd(choice, na.rm = TRUE),
.groups = "drop"
)

# By all three factors 
desc_3way <- df_long |>
group_by(target, group, block) |>
summarise(
n_trials = n(),
n_valid = sum(!is.na(choice)),
M = mean(choice, na.rm = TRUE),
SD = sd(choice, na.rm = TRUE),
.groups = "drop"
) |>
arrange(target, group, block)

desc_overall
desc_by_target
desc_by_group_block
desc_3way

# Save all tables 

saveRDS(desc_overall, p_processed("results", "desc_overall.rds"))
saveRDS(desc_by_target, p_processed("results", "desc_by_target.rds"))
saveRDS(desc_by_group_block, p_processed("results", "desc_by_group_block.rds"))
saveRDS(desc_3way, p_processed("results", "desc_3way.rds"))


```


## 3. Task Controls: Reward and Effort

```{r}

# Choice rates by reward level
desc_by_reward <- df_long |>
group_by(reward) |>
summarise(
n_trials = n(),
n_valid = sum(!is.na(choice)),
mean_choice = mean(choice, na.rm = TRUE),
sd_choice = sd(choice, na.rm = TRUE),
.groups = "drop"
)


# Choice rates by effort level
desc_by_effort <- df_long |>
group_by(effort) |>
summarise(
n_trials = n(),
n_valid = sum(!is.na(choice)),
mean_choice = mean(choice, na.rm = TRUE),
sd_choice = sd(choice, na.rm = TRUE),
.groups = "drop"
)

# Reward × Effort interaction pattern
desc_reward_effort <- df_long |>
group_by(reward, effort) |>
summarise(
n_trials = n(),
n_valid = sum(!is.na(choice)),
mean_choice = mean(choice, na.rm = TRUE),
sd_choice = sd(choice, na.rm = TRUE),
.groups = "drop"
)

desc_by_reward
desc_by_effort
desc_reward_effort

# Save all tables

saveRDS(desc_by_reward, p_processed("results", "desc_by_reward.rds"))
saveRDS(desc_by_effort, p_processed("results", "desc_by_effort.rds"))
saveRDS(desc_reward_effort, p_processed("results", "desc_reward_effort.rds"))
```

## 4. Moderator Variables

```{r}

sus_scores <- df_wide |>
select(starts_with("SUS_")) |>
mutate(across(everything(), as.numeric))

alpha_out <- psych::alpha(sus_scores, use = "pairwise")$total
omega_tot <- psych::omega(sus_scores, use = "pairwise")$omega.tot

SUS_total <- rowMeans(sus_scores, na.rm = TRUE)

sus_desc <- psych::describe(SUS_total)

alpha_out
omega_tot
sus_desc

saveRDS(sus_desc, p_processed("results", "sus_total_descriptives.rds"))



```

## 5. Ceiling Effects

```{r}


# By participant overall
participant_props <- df_long |>
group_by(ppn) |>
summarise(
n_trials = n(),
n_valid = sum(!is.na(choice)),
prop_high_effort = if_else(n_valid > 0, mean(choice, na.rm = TRUE), NA_real_),
.groups = "drop"
) |>
filter(!is.na(prop_high_effort))


ceiling_participants <- participant_props |>
summarise(
n_participants = n(),
at_90 = sum(prop_high_effort >= 0.90),
pct_90 = mean(prop_high_effort >= 0.90) * 100,
at_95 = sum(prop_high_effort >= 0.95),
pct_95 = mean(prop_high_effort >= 0.95) * 100,
at_100 = sum(prop_high_effort == 1.00),
pct_100 = mean(prop_high_effort == 1.00) * 100
)

ceiling_participants


# By participant and block
ceiling_by_block <- df_long |>
group_by(ppn, block) |>
summarise(
n_trials = n(),
n_valid = sum(!is.na(choice)),
prop_high_effort = if_else(n_valid > 0, mean(choice, na.rm = TRUE), NA_real_),
.groups = "drop"
) |>
filter(!is.na(prop_high_effort)) |>
group_by(block) |>
summarise(
n_participants = n(),
n_at_90 = sum(prop_high_effort >= 0.90),
pct_at_90 = mean(prop_high_effort >= 0.90) * 100,
n_at_95 = sum(prop_high_effort >= 0.95),
pct_at_95 = mean(prop_high_effort >= 0.95) * 100,
n_at_100 = sum(prop_high_effort == 1.00),
pct_at_100 = mean(prop_high_effort == 1.00) * 100,
.groups = "drop"
)

ceiling_by_block

# Distribution plot (discretize to 36 trials)

plot_data <- participant_props |>
mutate(prop_disc = round(prop_high_effort * 36) / 36) |>
count(prop_disc)

threshold <- 0.90

p_ceiling <- ggplot(plot_data, aes(x = prop_disc, y = n)) +
geom_col() +
scale_x_continuous(
breaks = plot_data$prop_disc,
labels = scales::percent_format(accuracy = 1)
) +
scale_y_continuous(
breaks = seq(0, max(plot_data$n), by = 10)
) +
geom_vline(xintercept = threshold, linetype = "dashed") +
annotate(
"text",
x = threshold,
y = max(plot_data$n) * 0.95,
label = "90% threshold",
hjust = -0.1
) +
labs(
x = "Proportion of high-effort choices (all 36 trials)",
y = "Number of participants"
) +
theme_minimal(base_size = 14) +
theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_ceiling

ggsave(
p_fig("descriptives", "figure1_ceiling_distribution.png"),
plot = p_ceiling,
width = 7, height = 5, dpi = 300
)


# Summary stats


ceiling_summary <- participant_props |>
summarise(
n_participants = n(),
mean = mean(prop_high_effort),
sd = sd(prop_high_effort),
median = median(prop_high_effort),
q25 = quantile(prop_high_effort, 0.25),
q75 = quantile(prop_high_effort, 0.75),
min = min(prop_high_effort),
max = max(prop_high_effort),
n_90 = sum(prop_high_effort >= 0.90),
pct_90 = mean(prop_high_effort >= 0.90) * 100,
n_95 = sum(prop_high_effort >= 0.95),
pct_95 = mean(prop_high_effort >= 0.95) * 100,
n_100 = sum(prop_high_effort == 1.00),
pct_100 = mean(prop_high_effort == 1.00) * 100
)

ceiling_summary
saveRDS(ceiling_summary, p_processed("results", "ceiling_summary.rds"))


# Within person variability (sd of choice)

within_person_variability <- df_long |>
group_by(ppn) |>
summarise(
n_valid = sum(!is.na(choice)),
sd_choice = if_else(n_valid > 1, sd(choice, na.rm = TRUE), NA_real_),
.groups = "drop"
) |>
filter(!is.na(sd_choice)) |>
summarise(
n_participants = n(),
mean_sd = mean(sd_choice),
median_sd = median(sd_choice),
n_zero_var = sum(sd_choice == 0),
pct_zero_var = mean(sd_choice == 0) * 100
)

within_person_variability
saveRDS(within_person_variability, p_processed("results", "within_person_variability.rds"))

# Ceiling by conditions

ceiling_by_group <- df_long |>
group_by(ppn, group) |>
summarise(
n_valid = sum(!is.na(choice)),
prop_high_effort = if_else(n_valid > 0, mean(choice, na.rm = TRUE), NA_real_),
.groups = "drop"
) |>
filter(!is.na(prop_high_effort)) |>
group_by(group) |>
summarise(
n_participants = n(),
n_at_90 = sum(prop_high_effort >= 0.90),
pct_at_90 = mean(prop_high_effort >= 0.90) * 100,
mean_prop = mean(prop_high_effort),
sd_prop = sd(prop_high_effort),
median_prop = median(prop_high_effort),
.groups = "drop"
)

ceiling_by_group
saveRDS(ceiling_by_group, p_processed("results", "ceiling_by_group.rds"))

```


# Visualizations

a)  Reward $\times$ Effort interaction

```{r}

plot_a_data <- df_long |> 
  group_by(reward, effort) |> 
  summarise(p = mean(choice == 1, na.rm = TRUE), .groups = "drop")


reward_effort_plot <- plot_a_data |>
  ggplot(aes(x = reward, y = p, color = effort, group = effort)) +
  geom_point(size = 3) +
  geom_line(size = 1) +
  scale_color_manual(values = c(
    "40%" = "#E15759",
    "90%" = "#4E79A7"
  )) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Reward",
    y = "Proportion of High-Effort Choices",
    color = "Effort"
  ) +
  theme_bw()



ggsave(
p_fig("descriptives", "reward_effort_plot.png"),
plot = reward_effort_plot,
width = 7, height = 5, dpi = 300
)

```

b)  Target $\times$ Block $\times$ Group

```{r}

plot_b_data <- df_long |>
  group_by(group, block, target) |>
  summarise(
    p_choice = mean(choice == 1, na.rm = TRUE),
    n       = sum(!is.na(choice)),
    .groups = "drop"
  )


target_block_group_plot <- plot_b_data |> 
  ggplot(aes(x = block,
           y = p_choice,
           color = target,
           group = target)) +
  geom_point(size = 3) +
  geom_line(size = 1) +
  facet_wrap(~ group, labeller = labeller(group = group_labels)) +
  labs(
    x     = "Block",
    y     = "Proportion of High-Effort Choices",
    color = "Target"
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_color_manual(values = pal_target) +
  theme_bw() +
  theme(
    legend.position = "top",
    legend.title    = element_blank()
  )

ggsave(
p_fig("descriptives", "target_block_group_plot.png"),
plot = target_block_group_plot,
width = 7, height = 5, dpi = 300
)


# Adjusting y axis to zoom in on the effect

zoomed_target_block_group_plot <- plot_b_data |> 
  ggplot(aes(x = block,
             y = p_choice,
             color = target,
             group = target)) +
  geom_point(size = 3) +
  geom_line(size = 1) +
  facet_wrap(~ group, labeller = labeller(group = group_labels)) +
  labs(
    x     = "Block",
    y     = "Proportion of High-Effort Choices",
    color = "Target"
  ) +
  scale_y_continuous(limits = c(0.85, 1), 
                     breaks = seq(0.85, 1, 0.025)) +
  scale_color_manual(values = pal_target) +
  theme_bw() +
  theme(
    legend.position = "top",
    legend.title    = element_blank()
  )

ggsave(
p_fig("descriptives", "zoomed_target_block_group_plot.png"),
plot = zoomed_target_block_group_plot,
width = 7, height = 5, dpi = 300
)

```
