---
title: "Missingness"
author: "Milena Costa"
date: "2025-11-17"
output: pdf_document
---

```{r setup, include=FALSE}

source(here::here("scripts","_common.R"), local = knitr::knit_global())
knitr::opts_chunk$set(echo = TRUE)  # override so this file shows code

df_wide <- read.csv(here::here("data","raw","data_2.csv"))


```

# Missing Data Analysis

## 1. Exploring missing data

```{r}


prop_miss(df_wide)

print(miss_var_summary(df_wide), n=50) # Missingness per variable

df_wide |> miss_case_summary() # Missingness per case
 
aggr(df_wide, numbers = TRUE, prop = FALSE) # missing pattern



missing_1 <- vis_miss(df_wide,
         sort_miss = TRUE)



ggsave("figures/missingness/missingness_plot.png", 
       missing_1, width = 7, height = 5, dpi = 300)

missing_2 <- gg_miss_var(df_wide)

ggsave("figures/missingness/missingness_plot2.png", 
       missing_2, width = 7, height = 5, dpi = 300)



df_nab <- df_wide |>  
  bind_shadow()


missing_3 <- df_nab |> 
  ggplot(aes(x = nom_like_NA, y = SUS_1)) +
  geom_boxplot()


ggsave("figures/missingness/missingness_plot3.png", 
       missing_3, width = 7, height = 5, dpi = 300)
```

# 2. Missing Data Mechanism


MAR: missingness associated with observed variables (but not the missing 
variable itself)
MCAR: missingness not associated with any variables (observed or unobserved)
MNAR: missingness associated with the unobserved (missing) values themselves

Here we make some boxplots with a few variables:


```{r}


vis_miss(df_wide, sort_miss = TRUE)


gg_miss_var(df_wide)


df_nab |> 
  ggplot(aes(x = nom_like_NA, y = SUS_1)) +
  geom_boxplot()


# --- nom_like --- #

df_nab |> 
  ggplot(aes(x = nom_like_NA,
             y = class)) +
  geom_boxplot()



df_nab |> 
  ggplot(aes(x = nom_like_NA,
             y = group)) +
  geom_boxplot()


df_nab |> 
  ggplot(aes(x = nom_like_NA,
             y = school)) +
  geom_boxplot()


plot_sus_na <- df_nab |> 
  ggplot(aes(x = nom_like_NA,
             y = SUS_1)) +
  geom_boxplot()


ggsave("figures/missingness/missingness_plot4.png", 
       plot_sus_na, width = 7, height = 5, dpi = 300)


# --- SUS scores --- #


df_nab |> 
  ggplot(aes(x = SUS_1_NA,
             y = class)) +
  geom_boxplot()



df_nab |> 
  ggplot(aes(x = SUS_1_NA,
             y = group)) +
  geom_boxplot()


df_nab |> 
  ggplot(aes(x = SUS_1_NA,
             y = school)) +
  geom_boxplot()


# --- X2 --- #


df_nab |> 
  ggplot(aes(x = X2_CLIMATE_2hard90_NA,
             y = class)) +
  geom_boxplot()



df_nab |> 
  ggplot(aes(x = X2_CLIMATE_2hard90_NA,
             y = group)) +
  geom_boxplot()


df_nab |> 
  ggplot(aes(x = X2_CLIMATE_2hard90_NA,
             y = school)) +
  geom_boxplot()




```

