---
title: "3_analysis"
author: "Milena Costa"
date: "2025-11-03"
output: pdf_document
---

```{r setup, include=FALSE}

source(here::here("scripts","_common.R"), local = knitr::knit_global())

df_long <- readRDS(p_processed("df_long.rds"))

ctrl <- lme4::glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e6))

```

# Statistical Analysis

## Research Question 1

## 1. Model Selection

### 1.1. Random Effects Structure

```{r}


# ---- Step 1: Test whether classroom random intercept is needed ---- #


# Baseline model (intercept only)

m0 <- glmer(
  choice ~ 1 + (1 | ppn),  
  data = df_long, 
  family = binomial, 
  control = ctrl 
)


m1_with_class <- glmer(
  choice ~ target * group * block + reward * effort + school +
    (1 | ppn) + (1 | class),
  data = df_long, 
  family = binomial, 
  control = ctrl 
)


m1_without_class <- glmer(
  choice ~ target * group * block + reward * effort + school +
    (1 | ppn),
  data = df_long, 
  family = binomial, 
  control = ctrl 
)


# Compare models
print(anova(m1_without_class, m1_with_class))

# Check for singularity

print(VarCorr(m1_with_class))

# ---- Step 2: Test whether random slopes for block is needed ---- #

m_rslopes <- glmer(
  choice ~ target*group*block + reward*effort + school +
    (1 + block | ppn),
  data = df_long, family = binomial, control = ctrl 
)

# Compare models

anova(m1_without_class, m_rslopes)

# Check for singularity

print(VarCorr(m_rslopes))

isSingular(m_rslopes)



```

Variance for class is estimated near-zero, resulting in singularity issues. LRT tests do not favor class random intercept. Therefore, model without class random intercept is retained.

Adding random slopes for block significantly improves the model fit. This indicates that participants differ substantially in how their choices evolve across blocks.

### 1.2. Interaction effects

```{r}

### Test 3-way interaction 

# Full model with three-way interaction
m_full <- glmer(
  choice ~ target * group * block + reward * effort + school +
    (1 + block | ppn),
  data = df_long, 
  family = binomial, 
  control = ctrl 
)

# Model without three-way interaction (only two-way)
m_no_3way <- glmer(
  choice ~ target + group + block + 
    target:group + target:block + group:block +  
    reward * effort + school +
    (1 + block | ppn),
  data = df_long, 
  family = binomial, 
  control = ctrl 
)



# Compare models
anova(m_no_3way, m_full)

```

The 3-way interaction is not significant. However, given our hypotheses, we do not remove any terms.

### 1.3. Final Model

The final model is given below.

```{r}

final_model <- glmer(
  choice ~ target * group * block + reward * effort + school +
    (1 + block | ppn),  
  data = df_long,
  family = binomial,
  control = ctrl
)



# Check singularity
isSingular(final_model)

# Print variance components

print(VarCorr(final_model))

# Compare with baseline

print(anova(m0, final_model))




```

## 2. Model Summary and Visualizations 

```{r}

# Model summary

print(summary(final_model))

# Odds Ratios with 95% CIs

# Extract coefficients and SEs
log_odds <- lme4::fixef(final_model)
se <- sqrt(diag(vcov(final_model)))

# Calculate ORs and CIs
or_table <- data.frame(
  Predictor = names(log_odds),
  OR = exp(log_odds),
  SE = se,
  CI_Lower = exp(log_odds - 1.96 * se),
  CI_Upper = exp(log_odds + 1.96 * se),
  p_value = summary(final_model)$coefficients[, "Pr(>|z|)"]
)



# Round for display
or_table_display <- or_table
or_table_display[, c("OR", "SE", "CI_Lower", "CI_Upper")] <- 
  round(or_table_display[, c("OR", "SE", "CI_Lower", "CI_Upper")], 3)
or_table_display$p_value <- round(or_table_display$p_value, 4)

print(or_table_display)


theme_set(theme_sjplot())
ordered_terms <- c(
  # CONTROL VARIABLES (bottom)
  "school2",
  "school1_loc2",
  "effort90%",
  "reward10 points",
  "reward6 points",
  "reward10 points:effort90%",
  "reward6 points:effort90%",
  
  # MAIN EFFECTS
  "targetprosocial",
  "targetclimate",
  "groupnegative_norm",
  "grouppositive_norm",
  "blockpost",
  
  # TWO-WAY INTERACTIONS 
  "targetprosocial:groupnegative_norm",
  "targetclimate:groupnegative_norm",
  "targetprosocial:grouppositive_norm",
  "targetclimate:grouppositive_norm",
  "targetprosocial:blockpost",
  "targetclimate:blockpost",
  "groupnegative_norm:blockpost",
  "grouppositive_norm:blockpost",
  
  # THREE-WAY INTERACTIONS (top)
  "targetprosocial:groupnegative_norm:blockpost",
  "targetprosocial:grouppositive_norm:blockpost",
  "targetclimate:groupnegative_norm:blockpost",
  "targetclimate:grouppositive_norm:blockpost"
)

# Create custom labels
custom_labels <- c(
  # Controls (bottom)
  "School 2",
  "School (Location 2)",
  "Effort (90%)",
  "Reward (10 pts)",
  "Reward (6 pts)",
  "Reward (10 pts) × Effort (90%)",
  "Reward (6 pts) × Effort (90%)",
  
  # Main effects
  "Target: Prosocial",
  "Target: Climate",
  "Negative norm",
  "Positive norm",
  "Block (Post)",
  
  # Two-way
  "Negative × Prosocial",
  "Negative × Climate",
  "Positive × Prosocial",
  "Positive × Climate",
  "Post × Prosocial",
  "Post × Climate",
  "Negative × Post",
  "Positive × Post",
  
  # Three-way (top)
  "Negative × Post × Prosocial",
  "Positive × Post × Prosocial",
  "Negative × Post × Climate",
  "Positive × Post × Climate"
)


forest_rq1 <- plot_model(final_model, 
                     type = "est",
                     show.intercept = FALSE,
                     title = "Fixed Effects: Social Norm Manipulation (RQ1)",
                     vline.color = "red",
                     terms = ordered_terms,
           axis.labels = custom_labels) +
  theme_bw()


ggsave(p_fig("regression-plots", "forest_rq1.png"),
       plot = forest_rq1, width = 8, height = 10, dpi = 300)

```


### Visualizations

```{r}

pred_rq1 <- predict_response(final_model, terms = c("block", "group", "target"))

plot_rq1 <- ggplot(pred_rq1, aes(x = x, y = predicted, colour = group, fill = group, group = group)) +
  geom_hline(yintercept = 0.90, linetype = "dashed", color = "red", alpha = 0.5) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) + 
  facet_wrap(~facet, ncol = 3) +
  coord_cartesian(ylim = c(0.85, 1.00)) +  
  theme_bw() +
  theme(
    legend.position = "right",
    strip.background = element_rect(fill = "white"),
    panel.grid.minor = element_blank()  
  ) +
  labs(
    title = "Target × Group × Block interaction", 
    y = "Predicted P(High-Effort Choice)",
    x = "Block",
    colour = "Group",
    fill = "Group"  # Need this for the ribbon
  ) + scale_colour_manual(
    values = c("control" = "#999999",        # Gray
               "positive_norm" = "#009E73",  # Green
               "negative_norm" = "#E69F00"), # Orange
    labels = c("control" = "Control",
               "positive_norm" = "Positive norm",
               "negative_norm" = "Negative norm")
  ) +
  scale_fill_manual(
    values = c("control" = "#999999",        # Gray
               "positive_norm" = "#009E73",  # Green
               "negative_norm" = "#E69F00"), # Orange
    labels = c("control" = "Control",
               "positive_norm" = "Positive norm",
               "negative_norm" = "Negative norm")
  )


plot_rq1

ggsave(p_fig("regression-plots", "plot_rq1.png"),
       plot = plot_rq1, width = 7, height = 5, dpi = 300)

```

## 3. Planned Contrasts

Because the hypothesis refers to difference-in-difference, we conduct planned contrasts.

```{r}

# Estimated marginal means
emm_clim <- emmeans(
  final_model,
  ~ block | group,
  at   = list(target = "climate"),
  type = "response"
)

# Post - pre change per group
delta_clim <- contrast(emm_clim, method = "revpairwise")
confint(delta_clim)  

# Compare changes between groups (planned contrasts)
did_clim <- contrast(
  delta_clim,
  method = list(
    "Positive vs Control"  = c(-1,  1,  0),  
    "Negative vs Control"  = c(-1,  0,  1),  
    "Positive vs Negative" = c( 0,  1, -1)
  ),
  by = "contrast",
  adjust = "holm"
)

summary(did_clim)

confint(did_clim)


```

## 4. Model Diagnostics and Performance

```{r}

#  Simulate DHARMa residuals

sim <- simulateResiduals(
  fittedModel = final_model,
  n = 1000,
  plot = FALSE
)


# Main diagnostic plots

plot(sim) # QQ + residuals vs fitted

plotQQunif(sim)  # QQ only
plotResiduals(sim) # residuals vs predicted only

# Statistical tests

testUniformity(sim)      # Should be non-significant
testDispersion(sim)      # Should be non-significant
testOutliers(sim)     # OK if significant for GLMMs, consider type="bootstrap" if significant

# Predictor-specific residual checks

mf <- model.frame(final_model) # because of NA values 


plotResiduals(sim, mf$target, main = "Residuals vs. Target")
plotResiduals(sim, mf$group, main = "Residuals vs. Group")
plotResiduals(sim, mf$block, main = "Residuals vs. Block")
plotResiduals(sim, mf$reward, main = "Residuals vs. Reward")
plotResiduals(sim, mf$effort, main = "Residuals vs. Effort")
plotResiduals(sim, mf$school, main = "Residuals vs. School")


# Grouped residuals (random-effects diagnostic)
sim_ppn <- recalculateResiduals(sim, group = mf$ppn)
plot(sim_ppn)
testDispersion(sim_ppn)


performance::icc(final_model)

performance::r2(final_model)

# Save diagnostic plots

w <- 7
h <- 5
dpi <- 300

# Plots 

png(p_fig("diagnostics", "dharma_main.png"), width = w, height = h, 
    units = "in", res = dpi)
plot(sim)
dev.off()


# Performance metrics 

# ICC
icc_val <- performance::icc(final_model)
icc_val


# R-squared
r2_vals <- performance::r2(final_model)
r2_vals



```


## Research Question 2

We extend the model from RQ1 and fit two moderation models. Note: glmmTMB is used given the increased model complexity.

```{r}

# Fit baseline equivalent model in glmmTMB
baseline_glmmTMB <- glmmTMB(
  choice ~ target * group * block + reward * effort + 
                             school +
                             (1 + block | ppn),
                            data = df_long,
                            family = binomial
                            )

summary(baseline_glmmTMB)
# summary(final_model) # Models are identical

```

## 1. Social Susceptibility Moderation (H4)

```{r}

m_suscept_final <- glmmTMB(
  choice ~ group * block * target * susceptibility_c + 
           reward * effort + school +
    (1 + block | ppn),  
  data = df_long, 
  family = binomial
)


summary(m_suscept_final)

sjPlot::tab_model(m_suscept_final)

drop1(m_suscept_final, test = "Chisq")

```

### 1.1. Planned Contrasts

```{r}

# pick susceptibility values (mean +- 1 SD)
sus_mean <- mean(df_long$susceptibility_c, na.rm = TRUE)
sus_sd   <- sd(df_long$susceptibility_c, na.rm = TRUE)
sus_vals <- c(low = sus_mean - sus_sd,
              mean = sus_mean,
              high = sus_mean + sus_sd)

# EMMs by block within each group x susceptibility (climate only)

emm_sus <- emmeans(
  m_suscept_final,
  ~ block | group * susceptibility_c,
  at = list(target = "climate",
            susceptibility_c = sus_vals),
  type = "response"
)

# post - pre within each group x susceptibility
delta_sus <- contrast(emm_sus, method = "revpairwise")
confint(delta_sus)

# planned contrasts of the post-pre change, within each susceptibility level
did_sus <- contrast(
  delta_sus,
  method = list(
    "Positive vs Control"  = c(-1,  1,  0),
    "Negative vs Control"  = c(-1,  0,  1),
    "Positive vs Negative" = c( 0,  1, -1)
  ),
  by = c("susceptibility_c", "contrast"),
  adjust = "holm"
)

confint(did_sus)
```

### 1.2. Diagnostics and Performance 

```{r}

#  Simulate DHARMa residuals

sim_sus <- simulateResiduals(
  fittedModel = m_suscept_final,
  n = 1000,
  plot = FALSE
)


# Main diagnostic plots

plot(sim_sus) # QQ + residuals vs fitted

plotQQunif(sim_sus)  # QQ only
plotResiduals(sim_sus) # residuals vs predicted only

# Statistical tests

testUniformity(sim_sus)      # Should be non-significant
testDispersion(sim_sus)      # Should be non-significant
testOutliers(sim_sus)     # OK if significant for GLMMs, consider type="bootstrap" if significant

# Predictor-specific residual checks

mf_sus <- model.frame(m_suscept_final) # because of NA values 


plotResiduals(sim_sus, mf_sus$target, main = "Residuals vs. Target")
plotResiduals(sim_sus, mf_sus$group, main = "Residuals vs. Group")
plotResiduals(sim_sus, mf_sus$block, main = "Residuals vs. Block")
plotResiduals(sim_sus, mf_sus$reward, main = "Residuals vs. Reward")
plotResiduals(sim_sus, mf_sus$effort, main = "Residuals vs. Effort")
plotResiduals(sim_sus, mf_sus$school, main = "Residuals vs. School")


# Grouped residuals (random-effects diagnostic)
sim_ppn_sus <- recalculateResiduals(sim_sus, group = mf_sus$ppn)
plot(sim_ppn_sus)
testDispersion(sim_ppn_sus)


# Save diagnostic plots

w <- 7
h <- 5
dpi <- 300

# Plots 

png(p_fig("diagnostics", "dharma_sus.png"), width = w, height = h, units = "in", res = dpi)
plot(sim_sus)
dev.off()


# Performance metrics

performance::icc(m_suscept_final)

performance::r2(m_suscept_final)



```

### 1.3. Model Coefficients Plot 

```{r}
ordered_terms_suscept <- c(

  # SUSCEPTIBILITY MAIN EFFECT
  "susceptibility_c",
  
  # TWO-WAY: Susceptibility interactions
  "targetprosocial:susceptibility_c",
  "targetclimate:susceptibility_c",
  "blockpost:susceptibility_c",
  "groupnegative_norm:susceptibility_c",
  "grouppositive_norm:susceptibility_c",
  
  
  # THREE-WAY: With susceptibility
  "blockpost:targetprosocial:susceptibility_c",
  "blockpost:targetclimate:susceptibility_c",
  "groupnegative_norm:targetprosocial:susceptibility_c",
  "grouppositive_norm:targetprosocial:susceptibility_c",
  "groupnegative_norm:targetclimate:susceptibility_c",
  "grouppositive_norm:targetclimate:susceptibility_c",
  "groupnegative_norm:blockpost:susceptibility_c",
  "grouppositive_norm:blockpost:susceptibility_c",
  
  # FOUR-WAY
  "groupnegative_norm:blockpost:targetprosocial:susceptibility_c",
  "grouppositive_norm:blockpost:targetprosocial:susceptibility_c",
  "groupnegative_norm:blockpost:targetclimate:susceptibility_c",
  "grouppositive_norm:blockpost:targetclimate:susceptibility_c"
)

# Custom labels - REVERSED order
custom_labels_suscept <- c(
  # Susceptibility main
  "Susceptibility",
  
  # Two-way: Susceptibility
  "Prosocial × Susceptibility",
  "Climate × Susceptibility",
  "Post × Susceptibility",
  "Negative × Susceptibility",
  "Positive × Susceptibility",
  
  # Three-way: With susceptibility
  "Post × Prosocial × Susceptibility",
  "Post × Climate × Susceptibility",
  "Negative × Prosocial × Susceptibility",
  "Positive × Prosocial × Susceptibility",
  "Negative × Climate × Susceptibility",
  "Positive × Climate × Susceptibility",
  "Negative × Post × Susceptibility",
  "Positive × Post × Susceptibility",
  
  # Four-way (H4) (top)
  "Negative × Post × Prosocial × Susceptibility",
  "Positive × Post × Prosocial × Susceptibility",
  "Negative × Post × Climate × Susceptibility",
  "Positive × Post × Climate × Susceptibility"
)


forest_suscept <- plot_model(m_suscept_final, 
           type = "est",
           show.intercept = FALSE,
           title = "Fixed Effects: Susceptibility Moderation (H4)",
           vline.color = "red",
           terms = ordered_terms_suscept,
           axis.labels = custom_labels_suscept) +
  theme_bw()


ggsave(p_fig("regression-plots", "forest_suscept.png"),
       plot = forest_suscept, width = 8, height = 10, dpi = 300)

```


## 2. Classroom Cohesion Moderation (H5)

```{r}

m_cohesion_cre <- glmmTMB(
  choice ~ group * block * target * cohesion_z + 
           reward * effort + school +
    (1 + block | ppn) + (1 | class),  # with class RE
  data = df_long, 
  family = binomial
)

# Check classroom random effect variance
VarCorr(m_cohesion_cre)

performance::check_singularity(m_cohesion_cre) # singularity issues 


m_cohesion_final <- glmmTMB(
  choice ~ group * block * target * cohesion_z + 
    reward * effort + school +
    (1 + block | ppn),  
  data = df_long, 
  family = binomial
)


summary(m_cohesion_final)

sjPlot::tab_model(m_cohesion_final)

drop1(m_cohesion_final, test = "Chisq")



```

### 2.1. Planned Contrasts

```{r}
coh_vals <- c(low = -1, mean = 0, high = 1)

emm_coh <- emmeans(
  m_cohesion_final,
  ~ block | group * cohesion_z,
  at = list(
    target = "climate",
    cohesion_z = coh_vals
  ),
  type = "response"
)


delta_coh <- contrast(emm_coh, method = "revpairwise")
confint(delta_coh)


did_coh <- contrast(
  delta_coh,
  method = list(
    "Positive vs Control"  = c(-1,  1,  0),
    "Negative vs Control"  = c(-1,  0,  1),
    "Positive vs Negative" = c( 0,  1, -1)
  ),
  by = c("cohesion_z", "contrast"),
  adjust = "holm"
)

confint(did_coh)

```


### 2.2. Diagnostics and Performance 

```{r}

#  Simulate DHARMa residuals

sim_coh <- simulateResiduals(
  fittedModel = m_cohesion_final,
  n = 1000,
  plot = FALSE
)


# Main diagnostic plots

plot(sim_coh) # QQ + residuals vs fitted

plotQQunif(sim_coh)  # QQ only
plotResiduals(sim_coh) # residuals vs predicted only

# Statistical tests

testUniformity(sim_coh)      # Should be non-significant
testDispersion(sim_coh)      # Should be non-significant
testOutliers(sim_coh)     # OK if significant for GLMMs, consider type="bootstrap" if significant

# Predictor-specific residual checks

mf_coh <- model.frame(m_cohesion_final) # because of NA values 



plotResiduals(sim_coh, mf_coh$target, main = "Residuals vs. Target")
plotResiduals(sim_coh, mf_coh$group,  main = "Residuals vs. Group")
plotResiduals(sim_coh, mf_coh$block, main = "Residuals vs. Block")
plotResiduals(sim_coh, mf_coh$reward, main = "Residuals vs. Reward")
plotResiduals(sim_coh, mf_coh$effort, main = "Residuals vs. Effort")
plotResiduals(sim_coh, mf_coh$school, main = "Residuals vs. School")


# Grouped residuals (random-effects diagnostic)
sim_ppn_coh <- recalculateResiduals(sim_coh, group = mf_coh$ppn)
plot(sim_ppn_coh)
testDispersion(sim_ppn_coh)

# Save diagnostic plots

w <- 7
h <- 5
dpi <- 300

# Plots 

png(p_fig("diagnostics", "dharma_coh.png"), width = w, height = h, units = "in", res = dpi)
plot(sim_coh)
dev.off()

# Performance metrics


performance::icc(m_cohesion_final)

performance::r2(m_cohesion_final)

```

### 2.3. Model Coefficients Plot

```{r}
# Cohesion model 
ordered_terms_cohesion <- c(
  
  # COHESION MAIN EFFECT
  "cohesion_z",
  
  # TWO-WAY: Cohesion interactions
  "targetprosocial:cohesion_z",
  "targetclimate:cohesion_z",
  "blockpost:cohesion_z",
  "groupnegative_norm:cohesion_z",
  "grouppositive_norm:cohesion_z",
  
  # THREE-WAY: With cohesion
  "blockpost:targetprosocial:cohesion_z",
  "blockpost:targetclimate:cohesion_z",
  "groupnegative_norm:targetprosocial:cohesion_z",
  "grouppositive_norm:targetprosocial:cohesion_z",
  "groupnegative_norm:targetclimate:cohesion_z",
  "grouppositive_norm:targetclimate:cohesion_z",
  "groupnegative_norm:blockpost:cohesion_z",
  "grouppositive_norm:blockpost:cohesion_z",
  
  # FOUR-WAY
  "groupnegative_norm:blockpost:targetprosocial:cohesion_z",
  "grouppositive_norm:blockpost:targetprosocial:cohesion_z",
  "groupnegative_norm:blockpost:targetclimate:cohesion_z",
  "grouppositive_norm:blockpost:targetclimate:cohesion_z"
)

# Custom labels - REVERSED order
custom_labels_cohesion <- c(
  
  # Cohesion main
  "Cohesion",
  
  # Two-way: Cohesion
  "Prosocial × Cohesion",
  "Climate × Cohesion",
  "Post × Cohesion",
  "Negative × Cohesion",
  "Positive × Cohesion",
  
  # Three-way: With cohesion
  "Post × Prosocial × Cohesion",
  "Post × Climate × Cohesion",
  "Negative × Prosocial × Cohesion",
  "Positive × Prosocial × Cohesion",
  "Negative × Climate × Cohesion",
  "Positive × Climate × Cohesion",
  "Negative × Post × Cohesion",
  "Positive × Post × Cohesion ",  
  
  # Four-way (H5) (top)
  "Negative × Post × Prosocial × Cohesion",
  "Positive × Post × Prosocial × Cohesion",
  "Negative × Post × Climate × Cohesion",
  "Positive × Post × Climate × Cohesion"
)



forest_cohesion <- plot_model(m_cohesion_final, 
           type = "est",
           show.intercept = FALSE,
           title = "Fixed Effects: Susceptibility Moderation (H4)",
           vline.color = "red",
           terms = ordered_terms_cohesion,
           axis.labels = custom_labels_cohesion) +
  theme_bw()


ggsave(p_fig("regression-plots", "forest_suscept.png"),
       plot = forest_cohesion, width = 8, height = 10, dpi = 300)

```

