---
title: "3_analysis"
author: "Milena Costa"
date: "2025-11-03"
output: pdf_document
---

```{r setup, include=FALSE}

source(here::here("scripts","_common.R"), local = knitr::knit_global())

df_long <- readRDS(p_processed("df_long.rds"))

ctrl <- lme4::glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e6))

```

# Statistical Analysis

## Research Question 1

## 1. Model Selection

Do exposure to social norms about environmental behavior influence children’s effortful pro-environmental choices?

### Step 1: Test whether classroom random intercept is needed

```{r}

# Baseline model (intercept only)

m0 <- glmer(
  choice ~ 1 + (1 | ppn),  
  data = df_long, 
  family = binomial, 
  control = ctrl 
)


m1_with_class <- glmer(
  choice ~ target * group * block + reward * effort + school +
    (1 | ppn) + (1 | class),
  data = df_long, 
  family = binomial, 
  control = ctrl 
)


m1_without_class <- glmer(
  choice ~ target * group * block + reward * effort + school +
    (1 | ppn),
  data = df_long, 
  family = binomial, 
  control = ctrl 
)


# Compare models
print(anova(m1_without_class, m1_with_class))

# Check for singularity

print(VarCorr(m1_with_class))


```

Decision: Use model without class random intercept.

### Step 2: Test whether random slopes for block is needed

```{r}

m_rslopes <- glmer(
  choice ~ target*group*block + reward*effort + school +
    (1 + block | ppn),
  data = df_long, family = binomial, control = ctrl 
)

# Compare models

anova(m1_without_class, m_rslopes)

# Check for singularity

print(VarCorr(m_rslopes))



```

Decision: Adding random slopes for block significantly improves the model fit This indicates that participants differ substantially in how their choices evolve across blocks.

### Step 3: Test 3-way interaction (Target $\times$ Group $\times$ Block)

```{r}

# Full model with three-way interaction
m_full <- glmer(
  choice ~ target * group * block + reward * effort + school +
    (1 + block | ppn),
  data = df_long, 
  family = binomial, 
  control = ctrl 
)

# Model without three-way interaction (only two-way)
m_no_3way <- glmer(
  choice ~ target + group + block + 
    target:group + target:block + group:block +  
    reward * effort + school +
    (1 + block | ppn),
  data = df_long, 
  family = binomial, 
  control = ctrl 
)



# Compare models
anova(m_no_3way, m_full)

```

### Final Model Selection for RQ1

The 3-way interaction is not significant, but given that the research question requires directly testing this hypothesis, we retain the full model.

```{r}

final_model <- glmer(
  choice ~ target * group * block + reward * effort + school +
    (1 + block | ppn),  
  data = df_long,
  family = binomial,
  control = ctrl
)



saveRDS(final_model, p_processed("models", "final_model_RQ1.rds")) # save model


# Check singularity
isSingular(final_model)

# Print variance components

print(VarCorr(final_model))

# Compare with baseline

print(anova(m0, final_model))


```

## 2. Model Summary and Effect Sizes

```{r}

# Model summary

print(summary(final_model))

# Odds Ratios with 95% CIs

# Extract coefficients and SEs
log_odds <- lme4::fixef(final_model)
se <- sqrt(diag(vcov(final_model)))

# Calculate ORs and CIs
or_table <- data.frame(
  Predictor = names(log_odds),
  OR = exp(log_odds),
  SE = se,
  CI_Lower = exp(log_odds - 1.96 * se),
  CI_Upper = exp(log_odds + 1.96 * se),
  p_value = summary(final_model)$coefficients[, "Pr(>|z|)"]
)



# Round for display
or_table_display <- or_table
or_table_display[, c("OR", "SE", "CI_Lower", "CI_Upper")] <- 
  round(or_table_display[, c("OR", "SE", "CI_Lower", "CI_Upper")], 3)
or_table_display$p_value <- round(or_table_display$p_value, 4)

print(or_table_display)

saveRDS(or_table_display, p_processed("results", "or_table_RQ1.rds"))


theme_set(theme_sjplot())
sjPlot::plot_model(final_model,
                   order.terms = c( 21, 22, 23, 24, # 3-way interactions 
                                    17, 18,           # Group × Block interactions 
                                    15, 16,           # Target × Block interactions 
                                    11, 12, 13, 14,   # Target × Group interactions 
                                    5,                # blockpost 
                                    3, 4,             # group main effects 
                                    1, 2,             # target main effects 
                                    6, 7, 8,          # reward and effort main effects 
                                    19, 20,           # reward × effort interactions 
                                    9, 10             # school effects 
                   ),            # school effects)
                   vline.color = "red")



ggsave(p_fig("regression-plots", "model-estimates.png"),
       width = 7, height = 5, dpi = 300)




```

## 3. Model Performance Metrics

```{r}

# ICC
icc_val <- performance::icc(final_model)
round(icc_val$ICC_adjusted, 3)

# R-squared
r2_vals <- performance::r2(final_model)
round(r2_vals$R2_marginal, 3)
round(r2_vals$R2_conditional, 3)

```

## 4. Post-hoc Analyses

Only in the case of a significant interaction. This was not the case.


```{r}

# 4.1 Reward × Effort

emm_re <- emmeans(final_model, ~ reward * effort, type = "response")
contrast(emm_re, "pairwise", by = "reward", adjust = "holm")
contrast(emm_re, "pairwise", by = "effort", adjust = "holm")


gg_re <- ggpredict(final_model, terms = c("reward", "effort"))
plot_re <- plot(gg_re) +
  labs(
    title = "Predicted Probability of High-Effort Choice",
    y = "P(High-Effort Choice)",
    x = "Reward (points)"
  ) +
  theme_bw() +
  scale_y_continuous(limits = c(0,1))


# 4. 2. Target $\times$ Group interaction (Exploratory)

# Estimated marginal means
emm_tg <- emmeans(final_model, ~ target * group, type = "response")
contrast(emm_tg, "pairwise", by = "group", adjust = "holm")
contrast(emm_tg, "pairwise", by = "target", adjust = "holm")


gg_tg <- ggpredict(final_model, terms = c("target", "group"))
plot_tg <- plot(gg_tg) +
  labs(
    title = "Target × Group Interaction (Exploratory)",
    subtitle = "Differential effects across targets, particularly for prosocial",
    y = "Predicted P(High-Effort Choice)",
    x = "Social Target",
    color = "Group"
  ) +
  theme_bw() +
  scale_y_continuous(limits = c(0,1))

# 4.3 Block × Group × Target (Primary hypothesis display)


emm_full <- emmeans(final_model, ~ target * group * block,
                    type = "response")
 
prepost <- contrast(
  emm_full,
  method = "pairwise",
  by = c("target", "group")
)

as.data.frame(prepost)

# Difference-in-differences: Compare change scores across groups

did <- contrast(
  emm_full,
  interaction = c("pairwise", "pairwise"),
  by = "target",
  adjust = "holm"
)
as.data.frame(did)

# Visualization showing ceiling effect

gg_3way <- ggpredict(final_model, terms = c("block", "group", "target"))

plot_3way <- plot(gg_3way) +
  labs(
    title = "Target × Group × Block: Primary Hypothesis Test",
    subtitle = "No differential effects detected (ceiling effects prevent detection)",
    y = "Predicted P(High-Effort Choice)",
    x = "Block",
    color = "Group"
  ) +
  facet_wrap(~ facet, ncol = 3) +
  theme_bw() +
  scale_y_continuous(limits = c(0.85, 1.0)) +  # Show the ceiling
  geom_hline(yintercept = 0.92, linetype = "dashed", color = "red", alpha = 0.5)

# save all

ggsave(p_fig("regression-plots", "reward_effort_interaction.png"),
       plot = plot_re, width = 7, height = 5, dpi = 300)

ggsave(p_fig("regression-plots", "target_group_interaction.png"),
       plot = plot_tg, width = 7, height = 5, dpi = 300)

ggsave(p_fig("regression-plots", "three_way_interaction_primary_hypothesis.png"),
       plot = plot_3way, width = 7, height = 5, dpi = 300)


```


## 5. Final Model Diagnostics

```{r}

#  Simulate DHARMa residuals

sim <- simulateResiduals(
  fittedModel = final_model,
  n = 1000,
  plot = FALSE
)


# Main diagnostic plots

plot(sim) # QQ + residuals vs fitted

plotQQunif(sim)  # QQ only
plotResiduals(sim) # residuals vs predicted only

# Statistical tests

testUniformity(sim)      # Should be non-significant
testDispersion(sim)      # Should be non-significant
testOutliers(sim)     # OK if significant for GLMMs, consider type="bootstrap" if significant

# Predictor-specific residual checks

mf <- model.frame(final_model) # because of NA values 


plotResiduals(sim, mf$target, main = "Residuals vs. Target")
plotResiduals(sim, mf$group, main = "Residuals vs. Group")
plotResiduals(sim, mf$block, main = "Residuals vs. Block")
plotResiduals(sim, mf$reward, main = "Residuals vs. Reward")
plotResiduals(sim, mf$effort, main = "Residuals vs. Effort")
plotResiduals(sim, mf$school, main = "Residuals vs. School")


# Grouped residuals (random-effects diagnostic)
sim_ppn <- recalculateResiduals(sim, group = mf$ppn)
plot(sim_ppn)
testDispersion(sim_ppn)


performance::icc(final_model)

performance::r2(final_model)

# Save diagnostic plots

w <- 7
h <- 5
dpi <- 300

# Plots 

png(p_fig("diagnostics", "dharma_main.png"), width = w, height = h, units = "in", res = dpi)
plot(sim)
dev.off()

png(p_fig("diagnostics", "dharma_random.png"), width = w, height = h, units = "in", res = dpi)
plot(sim_ppn)
dev.off()

png(p_fig("diagnostics", "lattice_random.png"), width = w, height = h, units = "in", res = dpi)
randoms1 <- ranef(final_model)
lattice::qqmath(randoms1)
dev.off()


```

## Research Question 2 (glmmTMB)

Are these effects moderated by (a) individual social susceptibility and (b) classroom cohesion?

Note: the following analyses make use of \texttt{glmmTMB} due to the complexity of the models.

We first verify if lme4 and glmmTMB produce equivalent results

```{r}
# Fit baseline equivalent model in glmmTMB
baseline_glmmTMB <- glmmTMB(
  choice ~ target * group * block + reward * effort + 
                             school +
                             (1 + block | ppn),
                            data = df_long,
                            family = binomial
                            )

summary(baseline_glmmTMB)
# summary(final_model) # Models are identical

```

### Social Susceptibility Moderation (H4)

```{r}

m_suscept_final <- glmmTMB(
  choice ~ group * block * target * susceptibility_c + 
           reward * effort + school +
    (1 + block | ppn),  # no class RE
  data = df_long, 
  family = binomial
)

saveRDS(m_suscept_final, p_processed("models", "m_suscept_final.rds"))

summary(m_suscept_final)
sjPlot::tab_model(m_suscept_final)


```


#### Model Diagnostics

```{r}

#  Simulate DHARMa residuals

sim_sus <- simulateResiduals(
  fittedModel = m_suscept_final,
  n = 1000,
  plot = FALSE
)


# Main diagnostic plots

plot(sim_sus) # QQ + residuals vs fitted

plotQQunif(sim_sus)  # QQ only
plotResiduals(sim_sus) # residuals vs predicted only

# Statistical tests

testUniformity(sim_sus)      # Should be non-significant
testDispersion(sim_sus)      # Should be non-significant
testOutliers(sim_sus)     # OK if significant for GLMMs, consider type="bootstrap" if significant

# Predictor-specific residual checks

mf_sus <- model.frame(m_suscept_final) # because of NA values 


plotResiduals(sim_sus, mf_sus$target, main = "Residuals vs. Target")
plotResiduals(sim_sus, mf_sus$group, main = "Residuals vs. Group")
plotResiduals(sim_sus, mf_sus$block, main = "Residuals vs. Block")
plotResiduals(sim_sus, mf_sus$reward, main = "Residuals vs. Reward")
plotResiduals(sim_sus, mf_sus$effort, main = "Residuals vs. Effort")
plotResiduals(sim_sus, mf_sus$school, main = "Residuals vs. School")


# Grouped residuals (random-effects diagnostic)
sim_ppn_sus <- recalculateResiduals(sim_sus, group = mf_sus$ppn)
plot(sim_ppn_sus)
testDispersion(sim_ppn_sus)


performance::icc(m_suscept_final)

performance::r2(m_suscept_final)


# Save diagnostic plots

w <- 7
h <- 5
dpi <- 300

# Plots 

png(p_fig("diagnostics", "dharma_sus.png"), width = w, height = h, units = "in", res = dpi)
plot(sim_sus)
dev.off()


```


### Classroom Cohesion Moderation (H5) 

#### Capped Directed Density


```{r}


m_cohesion_final <- glmmTMB(
  choice ~ group * block * target * cohesion_z + 
           reward * effort + school +
    (1 + block | ppn) + (1 | class),  # with class RE
  data = df_long, 
  family = binomial
)

# Check classroom random effect variance
VarCorr(m_cohesion_H5)

summary(m_cohesion_final)
sjPlot::tab_model(m_cohesion_final)



```

##### Model Diagnostics

```{r}

#  Simulate DHARMa residuals

sim_coh <- simulateResiduals(
  fittedModel = m_cohesion_final,
  n = 1000,
  plot = FALSE
)


# Main diagnostic plots

plot(sim_coh) # QQ + residuals vs fitted

plotQQunif(sim_coh)  # QQ only
plotResiduals(sim_coh) # residuals vs predicted only

# Statistical tests

testUniformity(sim_coh)      # Should be non-significant
testDispersion(sim_coh)      # Should be non-significant
testOutliers(sim_coh)     # OK if significant for GLMMs, consider type="bootstrap" if significant

# Predictor-specific residual checks

mf_coh <- model.frame(m_cohesion_final) # because of NA values 



plotResiduals(sim_coh, mf_coh$target, main = "Residuals vs. Target")
plotResiduals(sim_coh, mf_coh$group,  main = "Residuals vs. Group")
plotResiduals(sim_coh, mf_coh$block, main = "Residuals vs. Block")
plotResiduals(sim_coh, mf_coh$reward, main = "Residuals vs. Reward")
plotResiduals(sim_coh, mf_coh$effort, main = "Residuals vs. Effort")
plotResiduals(sim_coh, mf_coh$school, main = "Residuals vs. School")


# Grouped residuals (random-effects diagnostic)
sim_ppn_coh <- recalculateResiduals(sim_coh, group = mf_coh$ppn)
plot(sim_ppn_coh)
testDispersion(sim_ppn_coh)


performance::icc(m_cohesion_final)

performance::r2(m_cohesion_final)

# Save diagnostic plots

w <- 7
h <- 5
dpi <- 300

# Plots 

png(p_fig("diagnostics", "dharma_coh.png"), width = w, height = h, units = "in", res = dpi)
plot(sim_coh)
dev.off()



# Model estimates plots 

theme_set(theme_sjplot())

p_cohesion_final <- sjPlot::plot_model(
  m_cohesion_final,
  type = "est",   
  sort.est = FALSE,             
  show.intercept = FALSE,
  vline.color = "red"
)

p_cohesion_final

ggsave(p_fig("regression-plots", "m-cohesion-final.png"),
       plot = p_cohesion_final, width = 7, height = 5, dpi = 300)

theme_set(theme_sjplot())

terms_suscept_keep <- c(
  # 3-way interactions 
  "grouppositive_norm:blockpost:susceptibility_c",
  "groupnegative_norm:blockpost:susceptibility_c",
  
  # 2-way interactions 

  "grouppositive_norm:susceptibility_c",
  "groupnegative_norm:susceptibility_c",
  "blockpost:susceptibility_c",
  
  # main effects 

  "susceptibility_c"

)

p_suscept_final <- sjPlot::plot_model(
  m_suscept_final,
  type = "est",
  terms = terms_suscept_keep,
  sort.est = FALSE,
  show.intercept = FALSE,
  vline.color = "red"
)

ggsave(p_fig("regression-plots", "m-suscept-final.png"),
       plot = p_suscept_final, width = 7, height = 5, dpi = 300)



```


