---
title: "05_power_analysis"
author: "Milena Costa"
date: "2026-02-20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)

# eval = FALSE not to actually run the models given computational requirement 


```

# Power Analysis for Mixed Effects Models

This document describes how to conduct a simulation-based power analysis for the GLMM models used in this project, using the \texttt{simr} package in R.

The effects of interest are:

RQ1: 3-way interaction target $\times$ group $\times$ block

RQ2 (H4): 4-way interaction group $\times$ block $\times$ target $\times$ susceptibility_c

RQ2 (H5): 4-way interaction group $\times$ block $\times$ target $\times$ cohesion_z

## Requirements

```{r}

install.packages("simr")
library(simr)

```

The models must already be fitted before running power simulations. See 04_statistical_modelling.Rmd for model specifications.

## 1. Set the expected effect size

If testing power with hypothetical effect size, override the relevant coefficient manually:

```{r}
# Example: set log-odds effect for the critical 4-way interaction
# Small effect ~ 0.2, moderate ~ 0.3–0.5

fixef(m_suscept_final)["grouppositive_norm:blockpost:targetclimate:susceptibility_c"] <- 0.3
fixef(m_cohesion_final)["grouppositive_norm:blockpost:targetclimate:cohesion_z"] <- 0.3


```

## 2. Test power for the critical term

Since the models are already fitted on the data, the coefficients are already estimated, and thus the above manual setup is not needed. You can run powerSim() directly:

```{r}
# RQ1: 3-way interaction
powerSim(
  final_model,
  test = fixed("targetclimate:grouppositive_norm:blockpost"),
  nsim = 500
)

# RQ2 H4: Susceptibility 4-way
powerSim(
  m_suscept_final,
  test = fixed("grouppositive_norm:blockpost:targetclimate:susceptibility_c"),
  nsim = 500
)

# RQ2 H5: Cohesion 4-way
powerSim(
  m_cohesion_final,
  test = fixed("grouppositive_norm:blockpost:targetclimate:cohesion_z"),
  nsim = 500
)

```

## 3. Power curve across sample sizes

You can simulate what would happen if you had more participants than you actually collected. You can then compute power at multiple sample sizes to see whether your actual N was sufficient, or identify the N where power reaches 80%.

```{r}
# Extend to maximum N you want to test
model_ext <- extend(m_suscept_final, along = "ppn", n = 600)

# Power curve
pc <- powerCurve(
  model_ext,
  test = fixed("grouppositive_norm:blockpost:targetclimate:susceptibility_c"),
  along = "ppn",
  breaks = c(100, 200, 300, 400, 500, 600),
  nsim = 500
)

# Plot and print results
plot(pc)
print(pc)

```

Repeat for other models as needed.

### Interpreting results

The output reports estimated power (%) with 95% confidence intervals at each sample size. If your actual N already exceeds 80% power, report that. If it doesn't, it should be acknowledged as a limitation.

### Computation notes

These simulations are computationally intensive. It is wise to start with nsim = 100 for a rough estimate, then increase to nsim=500 for the final reported analysis. Consider running overnight.

Estimated runtimes:

| nsim | Approximate runtime |
|------|---------------------|
| 100  | ~10–20 min          |
| 200  | ~20–40 min          |
| 500  | ~1–2 hours          |


**References**

Green, P., & MacLeod, C. J. (2016). SIMR: An R package for power analysis of generalised linear mixed models by simulation. Methods in Ecology and Evolution, 7, 493–498. <https://doi.org/10.1111/2041-210X.12504>

Pargent, F., K. Koch, T., Kleine, A.-K., Lermer, E., & Gaube, S. (2024). A tutorial on tailored simulation-based sample-size planning for experimental designs with generalized linear mixed models. *Advances in Methods and Practices in Psychological Science*, 7(4). <https://doi.org/10.1177/25152459241287132>
